|

### 2019-12-20 FinalPaper.Rmd
Some more edits

---
title: |
  | \vspace{3.5in} \LARGE Predicting the Outcomes of Tennis Tournaments
subtitle: "A Monte Carlo Approach"
author: "Olivia Beck, Emma Lewis, Ryan Volkert"
date: "Decemeber 19, 2019"
output: pdf_document
header - input: 
  - \usepackage{float}
bibliography: bibliography.bib

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("funs.R")
library(knitr)
library(kableExtra)
library(tidyverse)
set.seed(400)
```

@@

For our report, we replicated the methods outlined in @Tennis for predicting the outcome of a tennis tournament. We used the empirical probability of each player winning a rally given they served to calculate theoretical probabilities that the player wins a game, a set, a match, and a tournament given the other players in the tournament.

@@

The probability of winning a game, a set, and a match in tennis are calculated based on the each players probability of winning a point on a serve. This paper and the reference paper,@Tennis both make the assumption that each serve is an independently and identically distributed (iid) random variable. This assumption has been made in other reports relating to the probability of winning a game in other racket sports and even though serves in tennis are not iid random variables in reality. For most purposes this is not an outrageous assumption as the divergence from iid is small.

A game in tennis is played with one player serving and is won by the first player to score four or more points and to be at least two points ahead of the other player. Player A can win a game against player B by a score of (4,0), (4,1) or (4,2), or else the score becomes (3,3) which is called "deuce".

@@

To win a match a player in woman's format a player must win two out of three sets and in men's format a player must win three out of five sets.

@@

Data for the 2002 Wimbledon for men's and women's as well as 2002 US Open for men's and women's was provided in @Tennis. Data for 2019 Wimbledon for men's and women's was found at @2019data.

@@

 $$p_A^R = \frac{\text{Points Won On Serve}}{\text{Points Served}}$$

@@

$$p_A^G = (p_A^R)^4[1+4q_A^R+10(q_A^R)^2] +20(p_A^Rq_A^R)^3(p_A^R)^2[1-2p_A^Rq_A^R]^{-1}$$

@@

$$p_A^S =\sum_{j=0}^4p_A^S(6,j) + p_A^S(7,5) + p_A^S(6,6)p_A^T$$

$$p_A^T = \sum_{j=0}^5p_A^T(7,j) +  p_A^T(6,6)p_A^Rq_B^R[1 - p_A^Rp_B^R - q_A^Rq_B^R]^{-1}$$

$$p_A^M = (p_A^S)^2 + 2(p_A^S)^2p_B^S$$

$$p_A^M = (p_A^S)^3 + 3(p_A^S)^3p_B^S + 6(p_A^S)^3(p_B^S)^2$$

$$p^{TC}_i =  \boldsymbol{P}_{ij}(\boldsymbol{P}_{ik}\boldsymbol{P}_{kl} + \boldsymbol{P}_{il}\boldsymbol{P}_{lk} ) $$
Note that $\sum_{\forall i} p^{TC}_i =1$. We will sample one  player from the 4 semifinalists with the probabilities calculated above to determine the tournament winner.

Where, $p_A^S(i,j)$ is defined recursively as:

@@

And,  $p_A^T(i,j)$ is defined recursively as:

@@

# Reproducability

First we will consider data from the 2002 US Open for both men's and women's. We want to reproduce the probability of winning a game. Table 1 compares the empirical probability of winning a game versus the theoretical probability of winning a game calculated in @Tennis versus the probability of winning a game that we calculated for the 4 semi finalists in this tournament. We can see that the vales that we calculated and the values calculated in @Tennis are the same. These values do differ slightly from the empirical probabilities. Table 2 outlines the same comparisons except for the men's 2002 US Open.

Next, we want to compare the probability of winning a tournament. @Tennis does not provide their predictions of tournament winners, only their probabilities compared to what actually happened in the tournament. The paper also only considers the 4 semifinalists when predicting the tournament, so we will as well.

```{r, echo = FALSE, fig.pos="H"}

@@

kable(TDWomen, format = "latex", booktabs = TRUE, caption = "Data for the Womens Semifinalists in the 2002 U.S. Open Tournament
")%>%
          kable_styling(latex_options = "scale_down")

@@

```{r, echo = FALSE, fig.pos="H"}

@@

kable(TDMen, format = "latex", booktabs = TRUE, caption = "Data for the Mens Semifinalists in the 2002 U.S. Open Tournament
")%>%
          kable_styling(latex_options = "scale_down")

@@

```{r, echo=FALSE, message=F, warning=F, fig.pos="H", fig.cap = "2002 Women's US Open"}
W02 <- read_csv('2002 US Open tournament- Women.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

ggplot(data.frame(T1)) +
  geom_col(aes(x=Var1, y=Freq/2000)) +
  xlab("Players") + 
  ylab("Number of Times Won") +
  ggtitle("2002 Women's US Open")

We fist look at 2002 US Open's Women's tournament for the 4 semi finalists. We calculate the probability each semi finalist has of winning the tournament, then we sample one player to win the tournament with the probabilities we calculated. Figure 1 shows the results from 2000 tournaments played with these 4 semi finalists. From our simulations we predict that S. Williams wins the tournament, and in fact she did win this tournament.

Similarly, we want to consider the 2002 US Open for the men's semifinalists. Using the same method outline above, we can see in Figure 2 that we predict P. Sampras to win the tournament, and in fact he did win this tournament.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.pos="H", fig.cap = "2002 Men's US Open"}
mOpen <- read_csv('2002 US Open- Men.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

ggplot(data.frame(T2)) +
  geom_col(aes(x=Var1, y=Freq/2000)) +
  xlab("Players") +
  ylab("Proportion of Times Won") +
  ggtitle("2002 Men's US Open")

Next, we consider data from the 2002 Wimbledon. The probability of winning a rally for the 4 semifinalists for both women's and men's can be found in Table 3. Figure 3 shows the results for the 2002 Women's Wimbledon. We predict that V. Williams wins, but in reality S. Williams had an upset win. Some speculations as to why we predicted incorrectly can be found in the Limitations section. Figure 4 shows the same methods as Figure 3, but for the men's semi finalists. We predict L. Hewitt to win the tournament, and in fact he does.

```{r, echo = FALSE, fig.pos="H"}

@@

```
```{r, message=F, warning=F, fig.pos="H", fig.cap = "2002 Women's Wimbledon"}
wWim02 <- read_csv('2002 Wimbledon -Women.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

ggplot(data.frame(T3)) +
  geom_col(aes(x=Var1, y=Freq/2000)) +
  xlab('Players') +
  ylab('Proportion of Times Won') +
  ggtitle("2002 Women's Wimbledon")
```
```{r, message=F, warning=F, fig.pos="H", fig.cap = "2002 Men's Wimbledon"}
mWim02 <- read_csv('2002 Wimbledon- Men.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

@@

For our extension, we ran the same method on data from the 2019 Wimbledon Tournament for both men's and women's. The data for probability of winning a rally can be found in Table 4. As seen in Figure 5, we predict that B. Strycoa wins the tournament, but S. Halep actually won the 2019 Women's Wimbledon. Similarly, in Figure 6 we predict that Rodger Federer and Rafeal Nadal essentially have an equal probability of winning the tournament, but it was Novak Djokovic who actually won the tournament. Again, the discrepancies between who won the tournament and what we predicted are discussed in the limitations section.  
```{r, echo =FALSE, warning=FALSE, fig.pos="H", message=FALSE}
#Women
wWim19 <- read_csv('2019 Womens Wimbledon.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

Tab<-cbind(Tab1, Tab2)
kable(Tab, format = "latex", booktabs = TRUE, caption = "Data for the Semifinalists in the 2019 Wimbledon Tournament
")%>%
          kable_styling(full_width = T)

```{r, warning=FALSE, echo=FALSE, message=F, fig.pos="H", fig.cap = "2019 Women's Wimbledon"}

T4 <- table(base::sample(wWim19$Name, 2000, prob = wWim19$pTC, replace = TRUE))

ggplot(data.frame(T4)) +
  geom_col(aes(x=Var1, y=Freq/2000)) +
  xlab('Players') +
  ylab('Proportion of Times Won') +
  ggtitle("2019 Women's Wimbledon")
```

```{r, warning=FALSE, echo=FALSE, message=F, fig.pos="H", fig.cap = "2019 Men's Wimbledon"}

@@

# References

>>>>

---
title: |
  | \vspace{3.5in} \LARGE Predicting the Outcomes of Tennis Tournaments
subtitle: "A Monte Carlo Approach"
author: "Olivia Beck, Emma Lewis, Ryan Volkert"
date: "Decemeber 19, 2019"
output: pdf_document
bibliography: bibliography.bib

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("funs.R")
library(knitr)
library(kableExtra)
library(tidyverse)
opts_chunk$set(fig.pos = 'H')
set.seed(400)
```

@@

This report replicates the methods outlined by @Tennis for predicting the outcome of a tennis tournament. We used the empirical probability of each player winning a rally given they served to calculate theoretical probabilities that the player wins a game, a set, a match, and a tournament. We first used the same data that @Tennis used for the 2002 Wimbledon and US Open tournaments. Our report then expands on the original study by doing the same analysis for the 2019 Wimbledon tournament.

@@

The probability of winning a game, a set, and a match in tennis are calculated based on each player's probability of winning a point on a serve. This paper and the reference paper both make the assumption that each serve is an independently and identically distributed (iid) random variable. This assumption has been made in reports relating to the probability of winning a game in other racket sports. It is important to note that serves in tennis are not iid random variables in reality. For most purposes this the divergence from iid is small, so we ignore the dependency.

A game in tennis is played with one player serving and is won by the first player to score four or more points and to be at least two points ahead of the other player. Player A can win a game against Player B by a score of (4,0), (4,1) or (4,2), or else the score becomes (3,3) which is called "deuce".

@@

To win a match in the women's format, a player must win two out of three sets and three out of five in the men's format.

@@

Data for the 2002 tournaments was provided by @Tennis. Data for the 2019 Wimbledon for was found at @2019data.

@@

\begin{equation}
p_A^R = \frac{\text{Points Won On Serve}}{\text{Points Served}}
\end{equation}

@@

\begin{equation}
p_A^G = (p_A^R)^4[1+4q_A^R+10(q_A^R)^2] +20(p_A^Rq_A^R)^3(p_A^R)^2[1-2p_A^Rq_A^R]^{-1}
\end{equation}

@@

\begin{equation}
p_A^S =\sum_{j=0}^4p_A^S(6,j) + p_A^S(7,5) + p_A^S(6,6)p_A^T
\end{equation}

$p_A^S(i,j)$ is defined recursively as:

@@

\begin{equation}
p_A^T = \sum_{j=0}^5p_A^T(7,j) +  p_A^T(6,6)p_A^Rq_B^R[1 - p_A^Rp_B^R - q_A^Rq_B^R]^{-1}
\end{equation}

$p_A^T(i,j)$ is defined recursively as:

@@

\begin{equation}
p_A^M = (p_A^S)^2 + 2(p_A^S)^2p_B^S
\end{equation}

\begin{equation}
p_A^M = (p_A^S)^3 + 3(p_A^S)^3p_B^S + 6(p_A^S)^3(p_B^S)^2
\end{equation}

\begin{equation}
p^{TC}_i =  \boldsymbol{P}_{ij}(\boldsymbol{P}_{ik}\boldsymbol{P}_{kl} + \boldsymbol{P}_{il}\boldsymbol{P}_{lk} )
\end{equation}

Note that $\sum_{\forall i} p^{TC}_i =1$. We will sample one player from the 4 semifinalists with the probabilities calculated above to determine the tournament winner.

# Reproducibility

First we will consider data from the 2002 US Open for both the men's and women's tournaments. We want to reproduce the probability of winning a game. Tables 1 and 2 compare the empirical probability of winning a game to the theoretical probability of winning a game calculated by @Tennis and the probability that we calculated for the four semi-finalists in each tournament. We can see that the values that we calculated and the values calculated by @Tennis are the same. These values do differ slightly from the empirical probabilities.

Next, we want to check the probability of winning a tournament. @Tennis do not provide their predictions of tournament winners. They also only consider the four semi-semifinalists in each tournament, so we will as well.

```{r, echo = FALSE}

@@

kable(TDWomen, format = "latex", booktabs = TRUE, caption = "Data for the Women's Semifinalists in the 2002 U.S. Open Tournament
")%>%
          kable_styling(latex_options = "scale_down")

@@

```{r, echo = FALSE}

@@

kable(TDMen, format = "latex", booktabs = TRUE, caption = "Data for the Men's Semifinalists in the 2002 U.S. Open Tournament
")%>%
          kable_styling(latex_options = "scale_down")

@@

```{r, echo=FALSE, message=F, warning=F, fig.cap = "2002 US Open", fig.show = 'hold', out.width = "50%"}
W02 <- read_csv('2002 US Open tournament- Women.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

ggplot(data.frame(T1)) +
  geom_col(aes(x=Var1, y=Freq/2000)) +
  xlab("Players") + 
  ylab("Proportion of Times Won") +
  ggtitle("2002 Women's US Open")

mOpen <- read_csv('2002 US Open- Men.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

ggplot(data.frame(T2)) +
  geom_col(aes(x=Var1, y=Freq/2000)) +
  xlab("Players") +
  ylab("Proportion of Times Won") +
  ggtitle("2002 Men's US Open")
```

We first look at the 2002 US Open. We calculated the probability each semi-finalist has of winning the tournament, then randomly sampled one player to win calculated weights. We repeated this 2,000 times, and took the player with the most sampled wins to be the predicted winner. Figure 1 shows the results from this method for the men's and women's tournaments. Our simulations predict that Serena Williams wins the women's tournament, and Pete Sampras wins the men's tournament. These results are consistent with the outcome of the actual 2002 US Open.

The 2002 Wimbledon is considered below.

```{r, echo = FALSE}

@@

```
```{r, message=F, warning=F, fig.cap = "2002 Wimbledon", fig.show = 'hold', out.width = "50%"}
wWim02 <- read_csv('2002 Wimbledon -Women.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

ggplot(data.frame(T3)) +
  geom_col(aes(x=Var1, y=Freq/2000)) +
  xlab('Players') +
  ylab('Proportion of Times Won') +
  ggtitle("2002 Women's Wimbledon")

mWim02 <- read_csv('2002 Wimbledon- Men.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

The probability of winning a rally for the four semifinalists for both women's and men's can be found in Table 3. Figure 2 shows the results of our method. We predict that Venus Williams wins, but in reality Serena Williams had an upset win. Some speculations as to why we predicted incorrectly can be found in the limitations section. For the men's semi finalists, we predict L. Hewitt to win the tournament, and in fact he does.

@@

For our extension, we ran the same method on data from the 2019 Wimbledon Tournament for both men's and women's. The data for probability of winning a rally can be found in Table 4. As seen in Figure 5, we predict that B. Strycoa wins the tournament, but S. Halep actually won the 2019 Women's Wimbledon. Similarly, in Figure 6 we predict that Rodger Federer and Rafeal Nadal essentially have an equal probability of winning the tournament, but it was Novak Djokovic who actually won the tournament. Again, the discrepancies between who won the tournament and what we predicted are discussed in the limitations section.

```{r, echo=F, warning=FALSE, message=FALSE}
#Women
wWim19 <- read_csv('2019 Womens Wimbledon.csv') %>%
  filter(!is.na(Name)) %>%
  mutate(pRally = NA)

@@

Tab<-cbind(Tab1, Tab2)

kable(Tab, format = "latex", booktabs = TRUE, caption = "Data for the Semifinalists in the 2019 Wimbledon Tournament
")%>%
          kable_styling(full_width = T)
```
```{r, warning=FALSE, echo=FALSE, message=F, fig.cap = "2019 Wimbledon", fig.show = 'hold', out.width = "50%"}
T4 <- table(base::sample(wWim19$Name, 2000, prob = wWim19$pTC, replace = TRUE))

ggplot(data.frame(T4)) +
  geom_col(aes(x=Var1, y=Freq/2000)) +
  xlab('Players') +
  ylab('Proportion of Times Won') +
  ggtitle("2019 Women's Wimbledon")

@@

\newpage
# References
